---
title: 'Homework 7: Functions and error handling'
author: ~
date: '2019-10-04'
slug: homework-7-functions-and-error-handling
categories: []
tags: []
---

# Functions

Create a function that does the following:

1. Accepts 3 arguments:
    * The first argument should be a data frame
    * The second and third argument should be the index number of columns in the data frame. These should default to the first and second column of the data frame, respectively.
2. Returns a vector that is the sum of the two columns specified in the arguments
3. Throws an informative error if:
    * The object passed in as the first argument is not a data frame
    * Either of the two columns are not numeric, so they can't be added
4. Auto-prints the result of the function, if it is not assigned to a value (just like most base R functions do).
    
So, as an example, your function should behave as follows:

```{r, include=FALSE}
my_fun <- function(df, index.1 = 1, index.2 = 2) {
  if(!is.data.frame(df)) {
    stop("The first argument is not a data frame")
  }
  if(!is.numeric(df[ ,index.1])) {
    stop("The second argument is not the index of a numeric column")
  }
  if(!is.numeric(df[ ,index.2])) {
    stop("The third argument is not the index of a numeric column")
  }
  
  col.1 <- df[ , index.1]
  col.2 <- df[ , index.2]
  col.1 + col.2
}
```

* Adds two columns of a data frame:
```{r}
result.1 <- my_fun(mtcars, index.1 = 3, index.2 = 4)
print(result.1)
```

* Default values of the second and third arguments are 1 and 2, respectively:
```{r}
result.2 <- my_fun(mtcars) 
print(result.2)
```

* Auto-prints when the result is not assigned to an object:
```{r}
my_fun(mtcars)
```

* Throws informative errors if the first argument is not a data frame, or if the second or third arguments are not numeric
```{r, error = TRUE}
my_fun(letters, 1, 2)
```

```{r, error=TRUE}
my_fun(iris, 4, 5)
```

**Hint**: Inside the function, you will want to use bracket subetting to access the columns of the data frame, e.g. `mtcars[ , 3]`
