---
title: 'Lecture notes: Join operations using dplyr'
author: ~
date: '2019-09-25'
slug: lecture-notes-join-operations-using-dplyr
categories: []
tags: []
---

# Resources

* The "two-table verbs" [vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/two-table.html) for **dplyr**.
    * **Note:** a vignette is a how-to guide for a package, and any reputable package will have one. A vignette complements the function documentation by demonstrating how you might use the package in real life, with concrete examples.
* The [relational data](https://r4ds.had.co.nz/relational-data.html) chapter in R for Data Science.

To learn about joining data frames, we'll use the data sets in the **nycflights13** package. The value of this data set is that `flights` is big: 336,776 rows!

```{r, message=FALSE}
library(nycflights13)
library(tidyverse)

# let's look at the data sets
glimpse(flights)
```

```{r}
glimpse(airlines)
```

Note that `flights` and `airlines` share a single variable, `carrier`. You can guess that we could add the carrier names to `flights` by joining the two data frames, by matching values of the `carrier` column in `flights` to the `carrier` column in `airlines`.

Let's get rid of a few of the columns of flight for clarity, and do that join. 

```{r}
flights2 <- flights %>%
  select(year:day, hour, origin, dest, tailnum, carrier)

flights2 %>%
  left_join(airlines) %>%
  glimpse()
```

We did not explicitly specify which column we should join by, but R has noticed that the two data frames share a single column, `carrier`, and has used that column to make the join.

Imagine that the two matching columns had different names, but the same information. We would have to specify which columns are common between the data frames.

```{r}
# Rename the 'carrier' column of flights to 'airline.code'
flights2 <- flights2 %>%
  rename(airline.code = carrier)

left_joined_data <- flights2 %>%
  left_join(airlines, by = c("airline.code" = "carrier"))
glimpse(left_joined_data)
```

As it happens, the set of "carriers" in `flights` and `airlines` is identical.
```{r}
unique(airlines$carrier)
```

```{r}
sort(unique(flights$carrier)) 
```

*(Side note: I used `sort()` to arrange the `flights` data into alphabetical order in order to make the comparison clearer. Note how we can call a bunch of functions in a row by placing them in nested parentheses: the above statement reads as "get the `carrier` vector from `flights`, find the unique values, and then sort them. However, even a few nested functions become pretty difficult to read. This is exactly the problem that piping solves.)*

Let's add a new, fake airline, ["Nashawn Wade Airlines"](https://en.wikipedia.org/wiki/Soul_Plane), code "NW". We'll add this to `airlines`, and look at the effect of different kinds of joins.

```{r}
airlines_mod <- rbind(airlines, list("NW", "Nashawn Wade Airlines"))

flights %>% 
  left_join(airlines_mod) %>%
  nrow()
```

Now try a right join: 
```{r}
flights %>% 
  right_join(airlines_mod) %>%
  nrow()
```

Aha! An extra row. Just to be sure, we'll look at the last row. This is the row that corresponds to carrier "NW":

```{r}
flights %>% 
  right_join(airlines_mod) %>%
  filter(carrier == "NW")
```

All the data are `NA`, except for `carrier` and `name`.

What would happen if you did a `full_join()`? How about an `anti_join()`


# Speed

There is a base R function to join data frames called `merge()`. It is **really** inefficient! Let's use the `system.time()` function to see how long each of these functions takes to execute. (This is a pretty rough-and-ready way to see how fast commands execute - the answers will depend on your computer and what else your computer is trying to do at the moment - but roughly speaking, the `User` column is the actual amount of time your computer spent executing the code).

```{r}
system.time({
  merge(flights, airlines)
})
```
```{r}
system.time({
  left_join(flights, airlines)
})
```

Wow! `merge()` usually takes more than a second, while `left_join()` is closer to 1/20 of a second! That's a big difference!
