---
title: 'Lecture notes: Subsetting and working with data frames'
author: ~
date: '2019-09-17'
slug: lecture-notes-subsetting-and-working-with-data-frames
categories: []
tags: []
---

# Resources

## Split-apply-combine

* [Original paper](https://www.jstatsoft.org/article/view/v040i01) (although note that the package that paper describes, **plyr**, has been replaced by the tidyverse package **dplyr**, which has totally different syntax.)
* Software Carpentry lessons:
    * [Split-apply-combine conceptual introduction](https://swcarpentry.github.io/swc-releases/2016.06/r-novice-gapminder/12-plyr/) 
    * [Implementation in **dplyr**](http://swcarpentry.github.io/r-novice-gapminder/13-dplyr/index.html)
    * Note that, for some reason, the first Software Carpentry lesson is in **plyr** and the second is is in **dplyr**. This really makes no sense, as **dplyr** is the replacement for **plyr** - you're not supposed to use them at the same time, and indeed loading them at the same time will cause some [really weird stuff](https://github.com/tidyverse/dplyr/issues/347) to happen. So don't do it.

## Factors

* [Software Carpentry](http://swcarpentry.github.io/r-novice-gapminder/04-data-structures-part1/index.html#factors)
* [Hands-On Programming with R](https://rstudio-education.github.io/hopr/r-objects.html#factors)
* [Advanced R](https://adv-r.hadley.nz/vectors-chap.html#s3-atomic-vectors)

## Subsetting
* [Software Carpentry](http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting/index.html)
* [HOPR on modifying values](https://rstudio-education.github.io/hopr/modify.html)
* [Advanced R, Chapter 4](https://adv-r.hadley.nz/subsetting.html)


# Split-apply-combine

The underlying idea behind the tidyverse is the "split-apply-combine" philosophy, as described by Hadley Wickham [here](https://www.jstatsoft.org/article/view/v040i01) 

The basic idea is simple: virtually every data analysis pipeline can be understood as consisting of the following steps:

1. **Split** the data into groups (for instance, sites, individual samples, experimental treatments, etc.)
2. **Apply** some analytical procedure. This could be as simple as measuring an average value, or it could be some complex operation that requires many lines of code.
3. **Combine** the results of the analytical procedure, to allow comparisons among groups.

The tidyverse encourages us to think about data analysis using this paradigm, and makes it quite easy to implement. We'll spend the next few weeks learning how to implement this.

The tidyverse is almost like a separate language - if you want, you could almost implement your whole data analytic pipeline using tidyverse syntax, without using any base R.

Almost, but not quite. Base R is still very useful in a lot of situations. Also, the tidyverse is relatively new, not everyone likes it better than base R, and there's a lot of code out there written in base R that you may want to reuse, study, or modify.

So today we're reviewing/expanding on some of the subsetting concepts we talked about at the start of the class. Then we'll move on to implementing the split-apply-combine philosophy in the tidyverse, and from there move out to many of the other cool things you can do in tidyverse.

# Factors

This is taken from the Software Carpentry gapminder [lesson 5](http://swcarpentry.github.io/r-novice-gapminder/05-data-structures-part2/index.html#factors).

```{r, message = FALSE}
# re-create the cats data frame from a few weeks ago
cats <- data.frame(
  coat = as.factor(c("calico", "black", "tabby", NA)), # note I'm forcing hte data type, just to be sure
  weight = c(2.1, 5.0, 3.2, 3.3),
  likes_string = as.integer(c(1, 0, 1, 1)),
  age = c(2, 3, 5, 9)
)
```

Factors are funny: they look like character vectors, but they are actually integer vectors where each integer carries a label. We call the set of labels the 'levels' of the factor. The levels are coded as "attributes" of the integer vector.

```{r, message = FALSE}
attr(cats$coat, "levels")
```

Note that we can't just append a new value to a factor, the way we could with a character vector. Instead, we need to first define the new level, which assigns the new label to a new integer in the factor's attributes. Then, we add the element to the factor itself. Even though say that we're adding `"tortioseshell"`, R is actually adding the integer 4, because the label "tortioseshell" is associated with the integer 4 in the factor.

Notice that we can add rows to a data frame using the `rbind()` function (i.e., "row bind"), if we define the new row as a list.

```{r}
# add a new level
levels(cats$coat) <- c(levels(cats$coat), "tortoiseshell") # First we define the new level
cats <- rbind(cats, list("tortoiseshell", 3.3, TRUE, 9)) # Second we add a new row, as a list, to the data frame
cats
```


We could just as well define the new row as a data frame, in which case we have to explicitly create the row names of the new data frame, which must match the row names of the old data frame precisely.

```{r, message=FALSE}
cats <- rbind(cats, data.frame(coat = "tortoiseshell", 
                               weight = 3.3,
                               likes_string = TRUE,
                               age = 9))

# cats$coat <- as.character(cats$coat)
```
Note that `as.numeric()` works for factors, but not for character vectors:
```{r, message=FALSE}
as.numeric(cats$coat)
```

```{r, message=FALSE}
as.numeric(c("calico", "black", "tabby", NA))
```

## Getting rid of factors

In general we want to avoid working with factors as possible. `as.character()` is great here:

```{r}
as.character(cats$coat)
```

If we use `read.csv()` (or its cousins `read.table()`, `read.delim()`, etc.) to read data from a text file, we will want to override the default parameter that reads columns containing only characters (aka 'strings') as a factor:
```{r}
gapminder <- read.csv("../../data/gapminder_data.csv", stringsAsFactors = FALSE) # Note that my directory structure is a bit different from yours because of the way I build a website using this code
str(gapminder)
```


# Subsetting

Elements of a data frame are *indexed* by their rows first, and then columns. A blank space indicates "all of the elements", So, `cats[4, ]` gives us all of the columns of the 4th row of `cats`. 
```{r}
cats[4, ]
```

We can remove rows using negative subscripts
```{r}
cats[-4, ]
```

And we can specify multiple rows or columns with a vector.
```{r}
cats <- cats[c(-4, -5), ]
```

One very useful bit of syntax is to specify a range of rows or columns using the `:` operator, which creates a vector of integers:

```{r}
1:3
```

This shows up in subsetting quite frequently:

```{r}
cats[ , 1:3]
```

## Subsetting with indices vs. booleans

There are three major ways to subset an object in R:

* Subsetting by index 
* Subsetting by name
* Subsetting by condition

(Note that [Advanced R](https://adv-r.hadley.nz/subsetting.html) breaks out 6 different ways to subset, and organizes those ways a little different to what I've done. Hadley is of course more precisely correct, but I think what I'm describing is a little easier to understand and covers most practical needs.)

### Subsetting by index

Above, we subsetted by index: if we want columns 1 through 3 of `cats`, we use brackets to ask for all the columns and the *column indices* 1 through 3.

```{r}
cats[ , 1:3]
```

### Subsetting by name

We could also ask for the elements by name. **This is often a better idea than subsetting by index**, especially with columns of data frames, because we can be sure of what we're getting - i.e., we don't have to worry about whether the columns got rearranged without us noticing:

```{r}
cats[ , c("weight", "likes_string", "age")]
```

This is similar to subsetting a column of a data frame using the `$` sign (e.g., `cats$weight`) but note that when we use `[]` we put the name of the element we want in quotation marks, and with `$` we don't.

### Subsetting by condition

Condition, here, means logicals, i.e., TRUE or FALSE. It is often the case that we want only the data that satisfies some condition. Here we pass a vector of TRUEs and FALSEs, and we get back all the elements that correspond to a TRUE.

```{r}
cats[ , c(FALSE, FALSE, TRUE, TRUE)]
```

Of course we almost never generate that vector of logicals manually - most often, we generate it using a boolean expression, i.e., an expression that returns either TRUE or FALSE. So for example,
```{r}
to.drop <- names(cats) %in% c("age", "likes_string")
cats[ , !to.drop]
```
Note that `!` means 'not'; it turns all the TRUEs to FALSEs and vice-versa.

Another common use would be to pick out all the rows that satisfy some condition. Say, for instance, we wanted all the rows for cats that weigh more than 3 pounds:

```{r}
cats[cats$weight > 3, ]
```

That works because `cats$weight > 3` returns a vector with as many elements as there are elements in `cats$weight`, with TRUE for elements that are greater than 3 and FALSE otherwise. 
