---
title: A very brief introduction to S3 classes
author: ~
date: '2019-10-09'
slug: a-very-brief-introduction-to-s3-classes
categories: []
tags: []
---

**DRAFT**: I've posted this without explaining how to write methods for new classes. I'll update shortly to include that.

# Resources

* This lecture was taken with minor changes from the Wickham [Chapter 13](https://adv-r.hadley.nz/s3.html) on S3 classes in Advanced R.

# Classes and methods

* There are two main reasons to use classes in R.
    * It allows you to create well-defined and structured data objects, so that you can be (reasonably) sure that an object of a given class will be structured in a specific way.
    * It allows you to define methods for your class, e.g. `print()` and `plot`.
* R has at least four different object-oriented systems. These are called, logically enough, S3, S4, Reference Classes, and R6. For most users, S3 is what you want to use.

## Introduction

The thing that makes an R object belong to a class is that it has a `class` attribute. This is really simple: it is a bit like saying that something becomes a dog once you put a label on it that says "dog". 

```{r, message=FALSE}
library(sloop)
f <- factor(c("a", "b", "c"))
typeof(f)
```

We can remove the class label with `unclass()`. When we do that, we notice that `print()` looks different.

```{r}
unclass(f)
```

A *generic* isa function that behaves differently depending on what class of object we have passed to it. We can tell whether a function is a generic using `sloop::ftype()`:

```{r}
ftype(print)
```

```{r}
ftype(unclass)
```

Generally methods can be identified because they have a `.` in the name, in the form of `generic.class()`, e.g. `print.data.frame()`. Here, `print` is a generic, and `print.data.frame()` is the function that is executed when we call `print` on a `data.frame`. Note, however, that lots of base R functions that are not methods also have `.` in the function name: for instance, `t.test()` is a generic.
```{r}
ftype(t.test)
```

```{r}
ftype(print.data.frame)
```

## Assigning a class to an object

We can assign a class to an object by setting the class attribute. This can be done directly when you create the object: 
```{r}
x <- structure(list(), class = "my_class")
class(x)
```

but it is perhaps more instructive to see what happens when we do it after creating the object:
```{r}
x <- list()
attr(x, "class")
```
It is NULL because we have not yet set any attributes. But we can set the attribute using `class()`
```{r}
class(x) <- "my_class"
attr(x, "class")
```
Note that this allows you to set a class object to something that doesn't make sense. 
```{r, results="asis", error=TRUE}
mod <- lm(log(mpg) ~ log(disp), data = mtcars)
class(mod)
```
```{r, error=TRUE}
class(mod) <- "Date"
print(mod)
```
Of course this doesn't work, because the code you would need to print a `Date` object won't work on an `lm` object.

## Constructor functions

Constructor functions create a valid instance of a class. They should always be named with the format `new_`, followed by the class of object they create. See other rules in the Advanced R chapter I'm ripping off here.
```{r}
new_Date <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "Date")
}

new_Date(c(-1, 0, 1))
```

```{r}
new_difftime <- function(x = double(), units = "secs") {
  stopifnot(is.double(x))
  units <- match.arg(units, c("secs", "mins", "hours", "days", "weeks"))

  structure(x,
    class = "difftime",
    units = units
  )
}

new_difftime(c(1, 10, 3600), "secs")
```

## Validators

Constructors perform extremely basic checks to see that input data is of a valid type. It is wise to write validator functions to ensure that everything about the object you have just created meets the requirements of its class.
```{r, error = TRUE}
new_factor <- function(x = integer(), levels = character()) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))

  structure(
    x,
    levels = levels,
    class = "factor"
  )
}

validate_factor <- function(x) {
  values <- unclass(x)
  levels <- attr(x, "levels")

  if (!all(!is.na(values) & values > 0)) {
    stop(
      "All `x` values must be non-missing and greater than zero",
      call. = FALSE
    )
  }

  if (length(levels) < max(values)) {
    stop(
      "There must be at least as many `levels` as possible values in `x`",
      call. = FALSE
    )
  }

  x
}

validate_factor(new_factor(1:5, "a"))
```
This validator function has given us a useful, interpretable error message.
